\PYGEOMETRY{}: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2020-11-13 public/paper/pyg4ometry.tex
final changes after my (stuart's) final reading

\section{FLUKA to GDML conversion}
FLUKA geometry is based upon a limited set of primitives (referred to as
\emph{bodies}) which can be combined using Boolean operations. A
\emph{zone} consists of one or more bodies or \emph{subzones} combined
using intersections and subtractions.  Zones may then be further combined
using union operations to form \emph{regions}, which is defined as the
union of one or more zones, as well as a material.

@@

Each FLUKA body is represented in \PYGEOMETRY{} with a corresponding class,
and in turn each class has a method that returns a GDML primitive solid and
a method that returns that solid's rotation and position such that it
matches its FLUKA equivalent.  The expansion, translation and transform
geometry directives are each folded into one or more of these three
methods.  The mapping of FLUKA bodies to GDML solids is shown in
Table~\ref{tab:Fluka2Geant4}.  It is worth noting that many of the FLUKA
bodies are infinite in extent, but are mapped to finite GDML solids.  The
translation of infinite bodies to equivalent finite solids is one of the
main and most involved steps in the conversion process.  This mapping is
possible because whilst FLUKA bodies can be infinite in extent, all zones
and regions must be finite.  Zones and regions are then composed by
instantiating these classes and adding body instances to them.  Each
\pyinline{Zone} and \pyinline{Region} instance can then return its
equivalent GDML Boolean solid.

The FLUKA CSG ASCII is parsed using an ANTLR4-generated parser,
producing an AST.  The resulting AST is then inspected sequentially
(\emph{walked}) to populate \pyinline{Region} instances with zones and bodies.
With the \pyinline{Region} instances populated they can then be manipulated
and translated into GDML.  The translation involves a number of special
steps to bridge the two disparate formats and ensure the resulting GDML is
well-formed and usable in Geant4.  Some of these steps are simple, for
example in FLUKA unions can be disconnected, but in Geant4 specifically
only MultiUnions can be disconnected.  Therefore, MultiUnions are used
throughout the converted geometry instead of the more conventional binary unions.
Other procedures are more involved and are discussed in the rest of this
section.

@@

\subsection{Infinite bodies}
The majority of bodies in FLUKA are infinite in extent, and fall
broadly into four categories, half-spaces, infinitely-long cylinders,
infinitely-long elliptical cylinders and quadric surfaces.
Translating these bodies to Geant4 requires generating the equivalent
finite solid whilst retaining the same final finite Boolean shape.
This is achieved with the use of axis-aligned bounding boxes (AABBs),
in which the FLUKA body is translated to a finite solid with with
dimensions slightly larger than the AABB.  The lengths of infinite
(elliptical) cylinders are reduced to finite equivalents with lengths
slightly greater than the bounding box.  Similarly, half-spaces are
reduced to boxes with one face acting as that of the half-space face,
and quadric surfaces are sampled only over the volume denoted by the
AABB.  Furthermore, the positions of these solids are moved as close
to the bounding box as possible whilst retaining an identical final
Boolean geometry.   Resizing and translating all bodies with respect
to the region bounding boxes ensures that an identical region geometry can
be generated, albeit from much smaller constituent bodies.

Generating these bounding boxes over which the bodies should be translated
to GDML involves first evaluating each region with very large \pyinline{Tub},
\pyinline{EllipticalTube} and \pyinline{Box} instances (by default
\SI{50}{\km} in length), such that they are effectively infinite for most
reasonable use cases.  \PYGEOMETRY{}'s CSG meshing is then used to generate
a mesh for each region from which the axis-aligned bounding box can be
extracted.  Each region is then evaluated a second time with respect to its
 bounding box, with all of its constituent infinite solids being
reduced in size as described above.  The implementation is described in
Algorithm~\ref{algo:redundant-halfspace}. This algorithm works robustly for
half-spaces, infinite circular cylinders and infinite elliptical cylinders
as the number of facets belonging to the mesh for such solids is
independent of the size of the solid, so generating the initial mesh from
arbitrarily large solids works well.  Generating a quadric surface over a very large
volume in space whilst retaining topological information is computationally
very expensive, so to resolve this the user must provide the approximate
axis-aligned bounding box of any region in which a quadric is used.

@@

  \KwData{FLUKA regions to be converted to GDML.}
  \KwResult{GDML solids equivalent to the FLUKA regions built from
    minimally-sized primitive solids}

@@

  $ G \longleftarrow \emptyset$ \;
  \For{body $b$ and AABB $a$ in $E$}{
    \tcp{Map the fluka bodies to minimal GDML solids by converting with the
    minimal AABB, $a$.}
    $G \longleftarrow $ \ToGDMLSolid{$b, a$}\;
  }
  \For{$r$ in regions to be converted}{
    build the corresponding GDML for $r$ from the set of minimal GDML
    primitives in $G$.\;
  }

  \label{algo:redundant-halfspace}
  \caption{The infinite-body minimisation algorithm employed in the
    conversion of Fluka to GDML.}
\end{algorithm}

@@

\subsection{Coplanar faces}
Coplanar faces present a problem during conversion where the faces of two
union components or that of two regions are perfectly coplanar in FLUKA.
Coplanar faces of this nature are ubiquitous in FLUKA and present no
problems to the operation of the program.  However, in Geant4 these will
generally result in tracking errors.  These must be handled robustly to
ensure the resulting geometry is usable.  Coplanar faces are resolved
automatically by slightly decreasing the size of every body that is used in
an intersection, and increasing the size of every body used in a
subtraction.  These rules are inverted for nested subtractions.  This
algorithm ensures that all coplanar faces are removed and replaced with
resolvable, small gaps between faces.  This approach works well for
guaranteeing well-formed geometry that is free from tracking errors in Geant4.

@@

FLUKA supports modular geometries with the use of the \fluka{lattice}
command. Figure~\ref{fig:lattice} demonstrates this capability. The
arbitrarily complex \emph{basic unit} can be defined once and used
multiple times by placing one or more empty \emph{lattice cells} with the
associated rototranslation from that lattice cell to the basic
unit. The lattice cells themselves will generally lack structure
and simply serve as a reference to the basic unit. The rototranslated
lattice cell must fully contain the basic unit and all of the regions
within it. When a particle steps into the lattice cell, the
rototranslation is applied to that particle and it is is taken into the
basic unit, and the simulation continues within the basic unit. Any
particles leaving the basic unit will be translated back to the
lattice cell with the inverted rototranslation. Up to two
levels of lattice nesting the are supported in FLUKA.

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=6cm]{./diagrams/lattice}
\caption{The lattice feature demonstrating a lattice cell
  referring to its basic unit with the rototranslation~$R_\textrm{trans}$.
  Any particle entering the cell will be transformed onto the basic unit
  with $R_\textrm{trans}$, and when leaving the basic unit, back to the cell
  with $R_\textrm{trans}^{-1}$.}
\label{fig:lattice}
\end{center}
\end{figure}
This feature is clearly analogous to the logical/physical volume feature in
Geant4, although it is more implicit as the contents of a given logical
volume are explicitly stated, whereas the contents of a lattice cell
are implied by the combination of the rototranslation and the locations of
its lattice cell and basic unit.

Translating the lattice construct into a logical volume (basic unit) with
many physical volumes (lattice cells) requires associating each lattice
cell with the full contents of its corresponding basic unit (typically more
than one region). This is achieved by meshing the complete FLUKA geometry
and then rototranslating the lattice cell mesh with its associated
rototranslation. By construction this will translate the lattice cell mesh
directly onto the full basic unit mesh. This transformed lattice cell mesh
can then be intersected with all other meshes in the scene.  Intersections
with the transformed lattice cell mesh determine the contents of the basic
unit.  Finally, the lattice cell can simply be replaced with a physical
volume containing the logical volumes determined in the previous step.
Thus a basic unit with one or more lattice cells can be translated into
Geant4's logical volume with one or more physical volumes. As has been
stated, FLUKA supports two levels of lattice nesting, but currently the
conversion to GDML supports only one. However, extending to an extra level
is simple in that it only involves an additional application of a
rototranslation matrix before checking for intersections with the lattice
cell mesh.  This capability will be added in a future release.

@@

Figure~\ref{fig:fluka-to-geant4-cup} shows a Faraday cup implemented in
FLUKA and accurately translated to GDML using \PYGEOMETRY{}.  Many of the
steps described above were directly applied in this model and all the
features are tested and demonstrated in the repository.  This set of
algorithms for bridging FLUKA with GDML covers a very broad range of
geometries, however a number of possible improvements for the future
remain.  For example, FLUKA regions can in general be disconnected, that is
to say the resulting set of points defined by that region may consist of
two disconnected subvolumes.  In FLUKA this typically manifests itself in
the form of disconnected unions, but it is also allowed for intersections
and subtractions to be disconnected.  In the case of translating
disconnected unions from FLUKA to GDML, this presents no difficulties as
the GDML \pyinline{MultiUnion} solid is explicitly allowed to be formed from
disconnected parts.  However, it is forbidden for \pyinline{Intersection} or
\pyinline{Subtraction} solids to be disconnected in this way and is
therefore necessary to account for this difference between the two codes.
One solution to this problem would involve detecting and separating
disconnected meshes, and then placing each separated component as a
separate \pyinline{TesselatedSolid} instance.  This decomposition has not
been implemented and is left as future work.

Furthermore, the quadric surface conversion can be improved by specialising
on some of the individual forms of the quadric.  Simply converting every
quadric into a TesselatedSolid comes at a potential performance cost in the
tracking, as well as usability in \PYGEOMETRY{} as the user must provide an
AABB for every single region featuring a quadric body.  In some cases
tessellation is unavoidable (for example a hyperbolic paraboloid), but
parabolic cylinders could for example be mapped to an \pyinline{ExtrudedSolid}.
This could provide both a performance improvement in the tracking time and
make \PYGEOMETRY{} easier to use as an AABB would not need to be provided
beforehand.  This has not been implemented as quadrics are relatively
rarely used, but where quadrics are used, it is often in the form of
parabolic cylinders (e.g. magnet pole tips) and this specialisation in
particular would be worth implementing.

@@

\begin{table}[hbt!]
\centering
\begin{tabular}{ l  l  } \hline
Geant4 solid			& FLUKA region construction		\\ \hline
Box					& +2 XYP + 2 XZP + 2 YZP 		\\
Tube					& +ZCC - 2 PLA -2 XYP - ZCC	 	\\
CutTube				& +ZCC - 4 PLA -ZCC			\\
Cone				& +TRC - TRC - 2 PLA 			\\
Para					& + 6 PLA						\\
Trd					& + 6 PLA						\\
Trap					& + 6 PLA						\\
Sphere				& +SPH - SPH  - 2 PLA - 2 TRC	\\
Orb					& +SPH						\\
Torus				& +$N$ ZCC  - $N$ PLA			\\
Polycone				& +$N$ TRC -2 PLA				\\
Polyhedra				& +$N$ PLA					\\
Eltube				& +ZEC  - 2 XYP				\\
Ellipsoid				& +ELL - 2 XYP		 			\\
Elcone				& +QUA - 2 XYP				\\
Paraboloid			& +QUA - 2 XYP				\\
Hype					& +QUA - QUA - 2 XYP			\\
Tet					& +4 PLA						\\
Xtru					& +$N$ PLA \\
TwistedBox			& +$N$ PLA					\\
TwistedTtap			& +$N$ PLA					\\
TwistedTrd			& +$N$ PLA				 	\\
TwistedTube			& +$N$ PLA					\\
Arb8					& +$N$ PLA					\\
Tessellated			& +$N$ PLA				 	\\
Union				& $R_1 \cup R_2$				\\
Subtraction			& $R_1 - R_2$					\\
Intersection			& $R_1 \cap R_2$				\\
MultiUnion			& $R_1 \cup R_2 \cup R_3 \cup$	\\ \hline
\end{tabular}
\label{tab:geant2fluka}
\caption{GDML/Geant4 solids and the mapping to FLUKA regions.}
\end{table}
Understanding the algebra of FLUKA CSG and how it relates to Geant4's
concepts of solids, logical volumes and physical volumes must be understood
to perform an accurate conversion.  For example, a Geant4 logical volume
solid may correspond to a FLUKA region consisting of many zones
(i.e. unions), e.g.,~$R_1 = +z_1\: | +z_2$.  If this logical volume is
placed as a physical volume inside some mother volume with its solid
perhaps consisting of multiple zones, e.g., $R_2= +z_3 \: |~+z_4$, then
subtracting a hole in the mother solid to make room for the daughter
results in an equation of the form
\begin{equation}
  \begin{aligned}
R_1 - R_2 	& = & (+z_1 \: | +z_2) - ( +z_3 \: | +z_4) 			\\
			& = & (+z_1 - z_3 - z_4) \;  | \; (+z_2 - z_3 - z_4)\,.
\label{eqn:setDiff}
  \end{aligned}
\end{equation}
Geant4 has Boolean solids associated with difference, union and intersection, so in addition to
Equation~\ref{eqn:setDiff}, both $R_1 \cup R_2$ and $R_1 \cap R_2$ are required in FLUKA notation, so
\begin{equation}
  \begin{aligned}
R_1 \cup R_2 	& = & (+z_1 \: | +z_2)  \cup ( +z_3 \: | +z_4) \\
			& = & +z_1 \: | +z_2 |  +z_3 \: | +z_4
\label{eqn:setUnion}
\end{aligned}
\end{equation}
and
\begin{equation}
  \begin{aligned}
    R_1 \cap R_2 	& = & (+z_1 \: | +z_2) \cap ( +z_3 \: | +z_4) \\
			& = & +z_1 +z_3  \; | +z_1 +z_4 \; | +z_2 +z_3 \; | +z_2 +z_4\,.
  \end{aligned}
\label{eqn:setIntersection}
\end{equation}

FLUKA (apart from the \fluka{lattice} directive) has no sense of a volume hierarchy. Each body is placed with
translation, rotation and expansion geometry directives in global coordinates. A transformation from world
coordinates to a physical volume is built up by recursively applying daughter volume transformations
and  this is used to place FLUKA bodies. This  in practice is very similar to the procedure to create the
VTK visualisation already described in Section~\ref{sec:visualisation}.

@@

Figure~\ref{fig:gdml-flair} shows an example conversion from GDML to FLUKA. The example is a
vacuum chamber with three ConFlat flange (CF) beam pipes connected to CF flanges. The top and bottom plates
have been removed to display the geometry more clearly. The model is formed of \verb|Box| and
\verb|Tubs| and Boolean operations of subtraction, intersection and union.

@@

\subsection{Disjunctive normal form and degenerate surfaces}
Typically FLUKA will decompose a region into disjunctive normal form (DNF)
as runtime, this normal form is characterised as the union of intersections and subtractions,
\begin{equation}
R = z_1 \; | \;z_2\;  | \; z_3 	\; | \; z_4 \dots
\end{equation}
Defining regions in terms of the DNF allows the rapid test of whether a point is inside that region. Testing each zone
$z_i$ of $R$ can terminate if a point is determined to be inside any of the zones $z_i$. In general
Equation~\ref{eqn:logicalSubtraction} does not have the form of a DNF. If there are many levels of logical-physical volume
placement, then recursive application of  Equation~\ref{eqn:logicalSubtraction} will create a nested set of
parentheses. There are some conditions where a general Boolean expression can yield an exponential
explosion of the final DNF. There are well known algorithms to convert logical expressions to its DNF.
\PYGEOMETRY{} can simplify parentheses from a region by creating a corresponding SymPy~\cite{10.7717/peerj-cs.103} Boolean
expression and using the \verb|to_dnf| method.  Further simplification of the CSG tree leverages \PYGEOMETRY{}'s
meshing capabilities combined with CSG pruning algorithms based
on~\cite{pruning}.  FLUKA by default will try to expand all regions to
their DNF at runtime, which inevitably can result in the sort of exponential explosion already mentioned.  Until version 4.0, if FLUKA identified
such an explosion, it would report an error and exit, thus making such models impossible to run.  As of version 4.0, however, this
expansion can be disabled, and the tracking algorithms will walk the CSG trees verbatim, at some tracking efficiency cost.  Therefore, most
output generated from the GDML to FLUKA conversion described here can only be used with FLUKA 4.0.

\subsection{Materials}
\PYGEOMETRY{} converts GDML/Geant4 materials to FLUKA \verb|MATERIAL| and \verb|COMPOUND| cards.
Geant4 has a class \cpinline{G4Material} to assign material state (density, physical state, temperature and pressure)
to a logical volume. \cpinline{G4Material} has two main constructors, the first where an atomic number is supplied and
the second is when \cpinline{G4Element} instances and relative atomic or mass abundances are provided. The {\em simple}
\cpinline{G4Material} is converted to a \fluka{material} card, whilst the {\em element} \cpinline{G4Material}  is converted to a
\fluka{compound} card. There are similar issues when converting \cpinline{G4Element} to FLUKA, as \cpinline{G4Element}
can either be simple, i.e. defined only by atomic number and mass, or composite and defined by an admixture of
relative abundances of \cpinline{G4Isotope}. A similar mapping is performed so that if a \cpinline{G4Element} is simple it is
directly converted to a FLUKA \fluka{MATERIAL} card, and the {\it isotope} \cpinline{G4Element} is converted to a \fluka{compound}
card. Geant4 also defines a set of standard materials~\cite{Geant4MaterialDB} or compounds from the US National Institute
of Standards and Technology (NIST), so a user can, for example, simply
specify the name \pyinline{G4_STAINLESS-STEEL}. \PYGEOMETRY{}  contains a matching database and creates the appropriate FLUKA
cards from these names during the conversion. This database is updated when
necessary by running a small Geant4 program to output the appropriate material data.

>>>>

\section{FLUKA to GDML conversion}
FLUKA geometry is based upon a limited set of primitives (referred to as
\emph{bodies}) which can be combined using Boolean operations. A
\emph{zone} consists of one or more bodies or \emph{subzones} combined
using intersections and subtractions.  Zones may then be further combined
using union operations to form \emph{regions}, which are defined as the
union of one or more zones, as well as a material.

@@

Each FLUKA body is represented in \PYGEOMETRY{} with a corresponding class,
and in turn each class has a method that returns a GDML primitive solid and
a method that returns that solid's rotation and position such that it
matches its FLUKA equivalent.  The expansion, translation and transform
geometry directives are each folded into one or more of these three
methods.  The mapping of FLUKA bodies to GDML solids is shown in
Table~\ref{tab:Fluka2Geant4}.  It is worth noting that many of the FLUKA
bodies are infinite in extent, but are mapped to finite GDML solids.  The
translation of infinite bodies to equivalent finite solids is one of the
main and most involved steps in the conversion process.  This mapping is
possible because whilst FLUKA bodies can be infinite in extent, all zones
and regions must be finite.  Zones and regions are then composed by
instantiating their respective classes and adding body instances to them.  Each
\pyinline{Zone} and \pyinline{Region} instance can then return its
equivalent GDML Boolean solid.

The FLUKA CSG ASCII is parsed using an ANTLR4-generated parser,
producing an AST.  The resulting AST is then inspected sequentially
(\emph{walked}) to populate \pyinline{Region} instances with zones and bodies.
With the \pyinline{Region} instances populated they can then be manipulated
and translated into GDML.  The translation involves a number of special
steps to bridge the two disparate formats and ensure the resulting GDML is
well-formed and usable in Geant4.  Some of these steps are simple, for
example in FLUKA unions can be disconnected, but in Geant4 specifically
only multi-unions can be disconnected.  Therefore, multi-unions are used
throughout the converted geometry instead of the more conventional binary unions.
Other procedures are more involved and are discussed in the rest of this
section.

@@

\subsection{Infinite bodies}
The majority of bodies in FLUKA are infinite in extent, and fall
broadly into four categories, half-spaces, infinitely-long cylinders,
infinitely-long elliptical cylinders and quadric surfaces.
Translating these bodies to Geant4 requires generating the equivalent
finite solid whilst retaining the same final finite Boolean shape.
This is achieved with the use of axis-aligned bounding boxes (AABBs),
in which the FLUKA body is translated to a finite solid with 
dimensions slightly larger than the AABB.  The lengths of infinite
(elliptical) cylinders are reduced to finite equivalents with lengths
slightly greater than the bounding box.  Similarly, half-spaces are
reduced to boxes with one face acting as that of the half-space face,
and quadric surfaces are sampled only over the volume denoted by the
AABB.  Furthermore, the positions of these solids are moved as close
to the bounding box as possible whilst retaining an identical final
Boolean geometry.   Resizing and translating all bodies with respect
to the region bounding boxes ensures that an identical region geometry can
be generated, albeit from much smaller constituent bodies.

Generating these bounding boxes over which the bodies should be translated
to GDML involves first evaluating each region with very large \pyinline{Tub},
\pyinline{EllipticalTube} and \pyinline{Box} instances (by default
\SI{50}{\km} in length), such that they are effectively infinite for most
reasonable use cases.  \PYGEOMETRY{}'s CSG meshing is then used to generate
a mesh for each region from which the axis-aligned bounding box can be
extracted.  Each region is then evaluated a second time with respect to its
 bounding box, with all of its constituent infinite solids being
reduced in size as described above.  The implementation is described in
Algorithm~\ref{algo:redundant-halfspace}. This algorithm works robustly for
half-spaces, infinite circular cylinders and infinite elliptical cylinders
as the number of facets belonging to the mesh for such solids is
independent of the size of the solid, so generating the initial mesh from
arbitrarily large solids works well.  Generating a quadric surface over a very large
volume in space whilst retaining topological information is computationally
very expensive, so to resolve this the user must provide the approximate
axis-aligned bounding box of any region in which a quadric is used to
constrain the size of the mesh only to where it is needed.

@@

  \KwData{FLUKA regions to be converted to GDML.}
  \KwResult{GDML solids equivalent to the FLUKA regions built from
    minimally-sized primitive solids.}

@@

  $ G \longleftarrow \emptyset$ \;
  \For{body $b$ and AABB $a$ in $E$}{
    \tcp{Map the FLUKA bodies to minimal GDML solids by converting with the
    minimal AABB, $a$.}
    $G \longleftarrow $ \ToGDMLSolid{$b, a$}\;
  }
  \For{$r$ in regions to be converted}{
    build the corresponding GDML for $r$ from the set of minimal GDML
    primitives in $G$.\;
  }

  \label{algo:redundant-halfspace}
  \caption{The infinite-body minimisation algorithm employed in the
    conversion of FLUKA to GDML.}
\end{algorithm}

@@

\subsection{Coplanar faces}
Coplanar faces present a problem during conversion where the faces of two
union components or that of two regions are perfectly coplanar in FLUKA.
Coplanar faces of this nature are ubiquitous in FLUKA and present no
problems to the operation of the program.  However, in Geant4 these will
generally result in tracking errors.  These must be handled robustly to
ensure the resulting geometry is usable in Geant4.  Coplanar faces are resolved
automatically by slightly decreasing the size of every body that is used in
an intersection, and increasing the size of every body used in a
subtraction.  These rules are inverted for nested subtractions.  This
algorithm ensures that all coplanar faces are removed and replaced with
resolvable, small gaps between faces.  This approach works well for
guaranteeing well-formed geometry that is free from tracking errors in Geant4.

@@

FLUKA supports modular geometries with the use of the \fluka{lattice}
command. Figure~\ref{fig:lattice} demonstrates this capability. The
arbitrarily complex \emph{basic unit} can be defined once and used
multiple times by placing one or more empty \emph{lattice cells} with the
associated rototranslation from that lattice cell to the basic
unit. The lattice cells themselves will generally lack structure
and simply serve as a reference to the basic unit. The rototranslated
lattice cell must fully contain the basic unit and all of the regions
within it. When a particle steps into the lattice cell, the
rototranslation is applied to that particle transporting it to the
basic unit, and the simulation continues. Any
particles leaving the basic unit will be transported back to the
lattice cell with the inverted rototranslation. Up to two
levels of lattice nesting are supported in FLUKA.

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=6cm]{./diagrams/lattice}
\caption{The lattice feature demonstrating a lattice cell
  referring to its basic unit with the rototranslation~$R_\textrm{trans}$.
  Any particle entering the cell will be transformed onto the basic unit
  with $R_\textrm{trans}$, and when leaving the basic unit, back to the cell
  with $R_\textrm{trans}^{-1}$.}
\label{fig:lattice}
\end{center}
\end{figure}
This feature is clearly analogous to the logical/physical volume feature in
Geant4, although it is more implicit as the contents of a given logical
volume are explicitly stated, whereas the contents of a lattice cell
are implied by the combination of the rototranslation and the global positions of
its lattice cell and basic unit.

Translating the lattice construct into a logical volume (basic unit) with
many physical volumes (lattice cells) requires associating each lattice
cell with the full contents of its corresponding basic unit (typically more
than one region). This is achieved by meshing the complete FLUKA geometry
and then rototranslating the lattice cell mesh with its associated
rototranslation. By construction this will translate the lattice cell mesh
directly onto the full basic unit mesh. This transformed lattice cell mesh
can then be intersected with all other meshes in the scene.  Intersections
with the transformed lattice cell mesh determine the contents of the basic
unit.  Finally, the contents of lattice cell can simply be replaced with a physical
volume containing the logical volumes determined in the previous step.
Thus a basic unit with one or more lattice cells can be translated into a
Geant4 logical volume with one or more physical volumes. As has been
stated, FLUKA supports two levels of lattice nesting, but currently the
conversion to GDML supports only one. However, extending to an extra level
is simple in that it only involves an additional application of a
rototranslation matrix before checking for intersections with the lattice
cell mesh.  This capability will be added in a future release.

@@

Figure~\ref{fig:fluka-to-geant4-cup} shows a Faraday cup implemented in
FLUKA and accurately translated to GDML using \PYGEOMETRY{}.  Many of the
steps described above were directly applied in this model and all the
features are tested and demonstrated in the repository.  This set of
algorithms for bridging FLUKA with GDML covers a very broad range of
geometries, however a number of possible improvements remain for the
future.
For example, FLUKA regions can in general be disconnected, that is
to say the resulting set of points defined by that region may consist of
two disconnected subvolumes.  In FLUKA this typically manifests itself in
the form of a disconnected unions, but it is also allowed for intersections
and subtractions to be disconnected.  In the case of translating
disconnected unions from FLUKA to GDML, this presents no difficulties as
the GDML multi-union solid is explicitly allowed to be formed from
disconnected parts.  However, it is forbidden for \pyinline{Intersection} or
\pyinline{Subtraction} solids to be disconnected in this way and is
therefore necessary to account for this difference between the two codes.
One solution to this problem would involve detecting and separating
disconnected meshes, and then placing each separated component as a
separate \pyinline{TesselatedSolid} instance.  This decomposition has not
been implemented and is left as future work.

Furthermore, the quadric surface conversion can be improved by specialising
on some of the individual forms of the quadric.  Simply converting every
quadric into a TesselatedSolid comes at a potential performance cost in the
tracking, as well as usability in \PYGEOMETRY{} as the user must provide an
AABB for every single region featuring a quadric body.  In some cases
tessellation is unavoidable (for example a hyperbolic paraboloid), but
parabolic cylinders, for example, could be mapped to \pyinline{ExtrudedSolid}
instances.
This could provide both a performance improvement in the tracking time and
make \PYGEOMETRY{} easier to use as an AABB would not need to be provided
beforehand.  This has not been implemented as quadrics are relatively
rarely used, but where quadrics are used, it is often in the form of
parabolic cylinders (e.g. magnet pole tips) and this specialisation in
particular would be worth implementing.

@@

\begin{table}[hbt!]
\centering
\begin{tabular}{ l  l  } \hline
Geant4 solid			& FLUKA region construction		\\ \hline
Box					& +2 XYP + 2 XZP + 2 YZP 		\\
Tube					& +ZCC - 2 PLA -2 XYP - ZCC	 	\\
CutTube				& +ZCC - 4 PLA -ZCC			\\
Cone				& +TRC - TRC - 2 PLA 			\\
Para					& + 6 PLA						\\
Trd					& + 6 PLA						\\
Trap					& + 6 PLA						\\
Sphere				& +SPH - SPH  - 2 PLA - 2 TRC	\\
Orb					& +SPH						\\
Torus				& +$N$ ZCC  - $N$ PLA			\\
Polycone				& +$N$ TRC -2 PLA				\\
Polyhedra				& +$N$ PLA					\\
Eltube				& +ZEC  - 2 XYP				\\
Ellipsoid				& +ELL - 2 XYP		 			\\
Elcone				& +QUA - 2 XYP				\\
Paraboloid			& +QUA - 2 XYP				\\
Hype					& +QUA - QUA - 2 XYP			\\
Tet					& +4 PLA						\\
Xtru					& +$N$ PLA \\
TwistedBox			& +$N$ PLA					\\
TwistedTtap			& +$N$ PLA					\\
TwistedTrd			& +$N$ PLA				 	\\
TwistedTube			& +$N$ PLA					\\
Arb8					& +$N$ PLA					\\
Tessellated			& +$N$ PLA				 	\\
Union				& $R_1 \cup R_2$				\\
Subtraction			& $R_1 - R_2$					\\
Intersection			& $R_1 \cap R_2$				\\
MultiUnion			& $R_1 \cup R_2 \cup R_3 \cup$	\\ \hline
\end{tabular}
\label{tab:geant2fluka}
\caption{GDML/Geant4 solids and the mapping to FLUKA regions.}
\end{table}
Understanding the algebra of FLUKA CSG and how it relates to Geant4's
concepts of solids, logical volumes and physical volumes is necessary
to perform an accurate conversion.  For example, a Geant4 logical volume
solid may correspond to a FLUKA region consisting of many zones
(i.e. unions), e.g.,~$R_1 = +z_1\: |~+z_2$.  If this logical volume is
placed as a physical volume inside some mother volume with its solid
perhaps consisting of multiple zones, e.g., $R_2= +z_3 \: |~+z_4$, then
subtracting a hole in the mother solid to make room for the daughter
results in an equation of the form
\begin{equation}
  \begin{aligned}
R_1 - R_2 	& = & (+z_1 \: | +z_2) - ( +z_3 \: | +z_4) 			\\
			& = & (+z_1 - z_3 - z_4) \;  | \; (+z_2 - z_3 - z_4)\,.
\label{eqn:setDiff}
  \end{aligned}
\end{equation}
Geant4 has Boolean solids associated with difference, union and intersection, so in addition to
Equation~\ref{eqn:setDiff}, both $R_1 \cup R_2$ and $R_1 \cap R_2$ are required in FLUKA notation, so
\begin{equation}
  \begin{aligned}
R_1 \cup R_2 	& = & (+z_1 \: | +z_2)  \cup ( +z_3 \: | +z_4) \\
			& = & +z_1 \: | +z_2 |  +z_3 \: | +z_4
\label{eqn:setUnion}
\end{aligned}
\end{equation}
and
\begin{equation}
  \begin{aligned}
    R_1 \cap R_2 	& = & (+z_1 \: | +z_2) \cap ( +z_3 \: | +z_4) \\
			& = & +z_1 +z_3  \; | +z_1 +z_4 \; | +z_2 +z_3 \; | +z_2 +z_4\,.
  \end{aligned}
\label{eqn:setIntersection}
\end{equation}

Apart from the lattice feature, FLUKA has no sense of a volume hierarchy. Each body is placed with
translation, rotation and expansion geometry directives in global coordinates. A transformation from world
coordinates to a physical volume is built up by recursively applying daughter volume transformations
and  this is used to place FLUKA bodies. This  in practice is very similar to the procedure to create the
VTK visualisation already described in Section~\ref{sec:visualisation}.

@@

Figure~\ref{fig:gdml-flair} shows an example conversion from GDML to FLUKA. The example is a
vacuum chamber with three ConFlat flange (CF) beam pipes connected to CF flanges. The top and bottom plates
have been removed to display the geometry more clearly. The model is formed of \verb|Box|,
\verb|Tubs| and Boolean operations of subtraction, intersection and union.

@@

\subsection{Disjunctive normal form and degenerate surfaces}
FLUKA will typically decompose a region into disjunctive normal form (DNF)
at runtime, this normal form is characterised as the union of intersections and subtractions,
\begin{equation}
R = z_1 \; | \;z_2\;  | \; z_3 	\; | \; z_4 \dots
\end{equation}
Defining regions in terms of the DNF allows the rapid test of whether a point is inside that region. Testing each zone
$z_i$ of $R$ can terminate if a point is determined to be inside any of the zones $z_i$. In general
Equation~\ref{eqn:logicalSubtraction} does not have the form of a DNF. If there are many levels of logical-physical volume
placement, then recursive application of  Equation~\ref{eqn:logicalSubtraction} will create a nested set of
parentheses. There are some conditions where a general Boolean expression can yield an exponential
explosion of the final DNF. There are well known algorithms to convert logical expressions to its DNF.
\PYGEOMETRY{} can simplify parentheses from a region by creating a corresponding SymPy~\cite{10.7717/peerj-cs.103} Boolean
expression and using the \verb|to_dnf| method.  Further simplification of the CSG tree leverages \PYGEOMETRY{}'s
meshing capabilities combined with CSG pruning algorithms based
on~\cite{pruning}.  FLUKA will by default try to expand all regions to
their DNF at runtime, which inevitably can result in the sort of exponential explosion already mentioned.  Until version 4.0, if FLUKA identified
such an explosion, it would the expansion, report an error and exit, thus making such models impossible to run.  As of version 4.0, however, this
expansion can be disabled, and the tracking algorithms will walk the CSG trees verbatim, albeit at some tracking efficiency cost.  Therefore, most
output generated from the GDML to FLUKA conversion described here can only be used with FLUKA 4.0.

\subsection{Materials}
\PYGEOMETRY{} converts GDML/Geant4 materials to FLUKA \verb|MATERIAL| and \verb|COMPOUND| cards.
Geant4 has a class \cpinline{G4Material} to assign material state (density, physical state, temperature and pressure)
to a logical volume. \cpinline{G4Material} has two main constructors, the first where an atomic number is supplied and
the second is when \cpinline{G4Element} instances and relative atomic or mass abundances are provided. The {\em simple}
\cpinline{G4Material} is converted to a \fluka{material} card, whilst the {\em element} \cpinline{G4Material}  is converted to a
\fluka{compound} card. There are similar issues when converting \cpinline{G4Element} to FLUKA, as \cpinline{G4Element}
can either be simple, i.e. defined only by atomic number and mass, or composite and defined by an admixture of
relative abundances of isotopes. A similar mapping is performed so that if a \cpinline{G4Element} is simple it is
directly converted to a FLUKA \fluka{MATERIAL} card, and the {\it isotope} \cpinline{G4Element} is converted to a \fluka{compound}
card. Geant4 also defines a set of standard materials~\cite{Geant4MaterialDB} or compounds from the US National Institute
of Standards and Technology (NIST), so a user can, for example, simply
specify the name \pyinline{G4_STAINLESS-STEEL}. \PYGEOMETRY{}  contains a matching database and creates the appropriate FLUKA
cards from these names during the conversion. This database is updated when
necessary by running a small Geant4 program to output the appropriate material data.

\subsection{Discussion}
Overall the conversion to FLUKA input format from GDML is quite advanced
and stable. Relatively large experimental simulations have been converted
from GDML to FLUKA and have been used to produce simulation results. The
conversion described still requires a user to understand how geometry is
specified in both Geant4 and FLUKA. For example, as has been stated, a
Geant4 mother volume will, when translated to a FLUKA region, have all of
its daughters subtracted from its respective FLUKA region. If those daughter volumes, when
translated to FLUKA, consist of combinations of many FLUKA bodies, the
mother region can become very complex.  This complex
region will generally not be in disjunctive normal form and is inefficient
when viewed in FLUKA's graphical user interface (GUI), flair, and used for
simulation in FLUKA.  Users who wish to minimise this should ensure that
there are not large numbers of daughter volumes at the same level in the
hierarchy, unless corresponding to solids which translate to simple
primitives such as boxes.  Complex daughter volumes consisting of solids
that are complex Booleans should be placed inside solids that map
to primitive FLUKA bodies.  This ensures that expanding to the DNF is
computationally tractable, and in doing do FLUKA's tracking will be faster.
This is not a drastic constraint on the conversion functionality, however,
as this is actually typically how one would implement geometries in FLUKA
regardless.

With the above in mind it is perhaps preferable for a user creating
geometry from scratch to use \PYGEOMETRY{} and these conversion tools
as part of the process.  This would allow a user to target multiple codes with
a single source description with minimal additional effort.  However, there
are still a  outstanding technical issues with the conversion, which are
discussed in the rest of this section.
